MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-05-10 latex/paper/sections/05-7series_architecture.tex
start packing

Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
Note how the Tile arrangement is columnar, where each column will have only one Tile type.

@@

\textbf{FFs} \quad
FFs are synthesized to facilitate synchronous event-driven signal assignment. 
For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
Figure \ref{fig:ff_synthesis} shows an example of where a FF is typically synthesized. 
The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (\texttt{CE}). 
\begin{itemize}
\item \texttt{FDCE} - DFF with CE and Asynchronous Clear
\item \texttt{FDPE} - DFF with CE and Asynchronous Preset
\item \texttt{FDSE} - DFF with CE and Synchronous Set
\item \texttt{FDRE} - DFF with CE and Synchronous Reset
\end{itemize}
In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.

@@

This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.

>>>>

Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
Note how the Tile arrangement is columnar such that column consists of only one Tile type.

@@

\textbf{FFs} \quad
FFs are synthesized to facilitate synchronous event-driven signal assignment. 
For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
Figure \ref{fig:ff_synthesis} shows an example of where a FF is typically synthesized. 
The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (\texttt{CE}). 
\begin{itemize}
\item \texttt{FDCE} - DFF with CE and Asynchronous Clear
\item \texttt{FDPE} - DFF with CE and Asynchronous Preset
\item \texttt{FDSE} - DFF with CE and Synchronous Set
\item \texttt{FDRE} - DFF with CE and Synchronous Reset
\end{itemize}
In a typical HDL design, the vast majority of FFs will be synthesized as \texttt{FDRE}s with the occasional \texttt{FDSE}, as it is generally good practice to keep FPGA designs synchronous. 
A FF BEL may also host a \texttt{LATCH} Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.

@@

This architectural context provides the necessary background for understanding how a placement algorithm should account for resource and placement constraints and optimize wirelength in modern FPGA architectures.

### 2025-05-10 latex/paper/sections/08-netlist.tex
start packing

RapidWright represents the logical netlist objects via the \texttt{edif} classes:
\begin{itemize}
    \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
    \item \texttt{EDIFNet}: Represents a logical net within an \texttt{EDIFNetlist}.
    \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
    \item \texttt{EDIFCell}: Represents a logical cell in an \texttt{EDIFNetlist}. 
    \item \texttt{EDIFCellInst}: Represents an instance of an \texttt{EDIFCell}. 
    \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
    \item \texttt{EDIFPort}: Represents a port on an \texttt{EDIFCell}. 
    \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
\end{itemize}

Using these classes and their associated methods, we can traverse the logical netlist (\texttt{EDIFNetlist}) and analyze or manipulate it as we see fit. 
A netlist can be easily extracted from a \texttt{.dcp} design checkpoint file and traversed like the following example:

>>>>

RapidWright represents the logical netlist objects via the \texttt{edif} classes:
\begin{itemize}
    \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
    \item \texttt{EDIFNet}: A logical net within an \texttt{EDIFNetlist}.
    \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
    \item \texttt{EDIFCell}: A logical cell in an \texttt{EDIFNetlist}. 
    \item \texttt{EDIFCellInst}: An instance of an \texttt{EDIFCell}. 
    \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
    \item \texttt{EDIFPort}: A port on an \texttt{EDIFCell}. 
    \item \texttt{EDIFPortInst}: An instance of a port on an \texttt{EDIFCellInst}. 
    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
\end{itemize}

Using these classes and their associated methods, we can traverse the logical netlist (\texttt{EDIFNetlist}) and analyze or manipulate it as we see fit. 
A netlist can be easily extracted from a \texttt{.dcp} design checkpoint file and traversed like shown in Listing \ref{lst:netlist_extract}. 
This is performed on the same design shown in figure \ref{fig:flat_netlist}.

### 2025-05-10 latex/paper/sections/10-simulated_annealing.tex
start packing

\section{Placement}
\end{multicols}
{
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
    \label{fig:substages}
}
\begin{multicols}{2}
\label{sec:simulated_annealing}
With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures of \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.

@@

The first step in our placement flow is \textbf{prepacking}. 
Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
The job of the prepacker is to traverse the raw EDIF netlist, detect these cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.

@@

The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
It does not report the presence of any multi-cell structures (\texttt{CARRY4} chains, etc.). 
Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.

The code snippet \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
We first collect the cells in the design that are of type \texttt{CARRY4}, then iteratively traverse their Carry-Out (\texttt{CO}) to Carry-In (\texttt{CI}) nets to find incident \texttt{CARRY4} cells.
Each \texttt{CARRY4} chain has an anchor cell and a tail cell.
The anchor is found when the \texttt{CI} net connects to Ground (\texttt{GND}), while the tail is found when the \texttt{CO} port is null.

Similarly, \texttt{DSP48E1} cascades can be found and collected by traversing the \texttt{PCOUT} \texttt{ACOUT} and \texttt{BCOUT} nets.
\texttt{LUT-FF} pairs can be found by inspecting the LUT output (O) net and checking for FF input (DI) ports. 
We detect and collect these cell structures and consolidate them into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.

@@

    // Now we have the chain anchor as currentCell.
    // Now traverse in the opposite direction to find the chain tail.
    // Tail is found when the CO Port is null.
    // Collect the chain cells into an ordred list.
    List<EDIFCellInst> currentChain = new ArrayList<>();
    currentChain.add(currentCell);
    while (true) {
        System.out.println("while 2");
        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
        if (sourcePort == null) {
            // Found this chain's tail!
            break;
        }
        EDIFNet net = sourcePort.getNet();
        List<EDIFPortInst> netPorts = net.getPortInsts();
        for (EDIFPortInst netPort : netPorts) {
            EDIFCellInst sinkCell = netPort.getCellInst();
            if (netPort.getName().equals("CI") &&
                    sinkCell.getCellName().equals("CARRY4")) {
                currentCell = sinkCell;
                // Add the cell to the chain list.
                currentChain.add(currentCell);
                break;
            }
        }
    }
    System.out.println(currentChain.size());
    // Add currentChain to the list of chains
    carryChains.add(currentChain);
    // Remove currentChain from the list of cells
    carryCells.removeAll(currentChain);
} // end while()
return carryChains;
\end{lstlisting}

\subsection{Packing}
    \label{subsec:packing}

\subsection{Placement}
    \label{subsec:placement}
    Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
    This is where simulated annealing actually begins.

>>>>

\section{Placement}
\end{multicols}
{
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
    \label{fig:substages}
}
\begin{multicols}{2}
\label{sec:simulated_annealing}
With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures around \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.

@@

The first step in our placement flow is \textbf{prepacking}. 
Recall from the 7-Series architecture that there are certain multi-cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
The job of the prepacker is to traverse the raw EDIF netlist, detect these multi-cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.

@@

The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
It does not report the presence of any multi-cell structures (\texttt{CARRY4} chains, etc.). 
Thus, we must traverse the netlist to detect these multi-cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.

The code snippet in \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
We first collect the cells in the design that are of type \texttt{CARRY4}, then iteratively traverse their Carry-Out (\texttt{CO}) to Carry-In (\texttt{CI}) nets to find incident \texttt{CARRY4} cells.
Each \texttt{CARRY4} chain has an anchor cell and a tail cell.
The anchor is found when the \texttt{CI} net connects to Ground (\texttt{GND}), while the tail is found when the \texttt{CO} port is null. 
We can further detect if there are \texttt{LUT}s or \texttt{FF}s connected to the \texttt{CARRY4} cell and store that information in a data structure we will call \texttt{CarryCellGroup} as defined in figure \ref{fig:substages}.
This will help us in knowing which cells can be packed together into the same \texttt{Site} in the subsequent stages.

Similarly, \texttt{DSP48E1} cascades can be found and collected by traversing the \texttt{PCOUT} \texttt{ACOUT} and \texttt{BCOUT} nets.
\texttt{LUT-FF} pairs can be found by inspecting the LUT output (O) net and checking for FF input (DI) ports. 
We can bucket these LUT-FF pairs by finding the set of unique \texttt{CE} \texttt{SR} net pairs to know which group of LUT-FF pairs can be placed within the same \texttt{Site}. 
We detect and collect these multi-cellular structures and consolidate them into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.

@@

    // Now we have the chain anchor as currentCell.
    // Now traverse in the opposite direction to find the chain tail.
    // Tail is found when the CO Port is null.
    // Collect the chain cells into an ordred list.
    List<EDIFCellInst> currentChain = new ArrayList<>();
    currentChain.add(currentCell);
    while (true) {
        System.out.println("while 2");
        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
        if (sourcePort == null) {
            // Found this chain's tail!
            break;
        }
        EDIFNet net = sourcePort.getNet();
        List<EDIFPortInst> netPorts = net.getPortInsts();
        for (EDIFPortInst netPort : netPorts) {
            EDIFCellInst sinkCell = netPort.getCellInst();
            if (netPort.getName().equals("CI") &&
                    sinkCell.getCellName().equals("CARRY4")) {
                currentCell = sinkCell;
                // Add the cell to the chain list.
                currentChain.add(currentCell);
                break;
            }
        }
    }
    System.out.println(currentChain.size());
    // Add currentChain to the list of chains
    carryChains.add(currentChain);
    // Remove currentChain from the list of cells
    carryCells.removeAll(currentChain);
} // end while()

// Print out the carry chains. 
for (List<EDIFCellInst> chain : chains) {
    for (int i = 0; i < chain.size(); i++) {
        EDIFCellInst carry = chain.get(i);
        if (i == 0) {
            writer.write("\nAnchor Cell: " + carry.getName() + 
                ", CellType: " + carry.getCellName());
        } else {
            writer.write("\n\tCell: " + carry.getName() + 
                ", CellType: " + carry.getCellName());
        }
    }
}
\end{lstlisting}

\subsection{Packing}
\label{subsec:packing}
Now that we have our \texttt{PrepackedDesign} object keeping track of multi-cell structures on the \texttt{edif} level, we can start packing them into \texttt{SiteInst} objects on the \texttt{design} level. 
A \texttt{SiteInst} object can be created and its BELs populated by existing \texttt{EDIFHierCellInst} objects. 
A \texttt{SiteInst} is necessarily placed on a specific \texttt{Site} upon creation. 
Thus, the packer necessarily gives all \texttt{SiteInst}s an initial placement before randomization via simulated annealing in the following placement stage. 
In packing, will simply assign \texttt{SiteInst}s pseudorandomly onto device \texttt{Site}s on a first-come-first-serve basis by coordinate order (first SiteInst onto \texttt{SLICEL\_X0Y0}, second \texttt{SiteInst} onto \texttt{SLICEL\_X0Y1}, etc.).

\end{multicols}
\begin{lstlisting}[caption=\texttt{SiteInst} construction and BEL population.]
Constructor: 
    SiteInst(String name, Design design, SiteTypeEnum type, Site site)

Usage:
    SiteInst si = new SiteInst(<name>, <design>, <type>, <site>);
    si.createCell(<EDIFHierCellInst>, <BEL>);
\end{lstlisting}

\begin{lstlisting}[language=java, caption={Packing DSPCascades into \texttt{DSP48E1} \texttt{SiteInst}s.}, label={lst:carry_chains}]
private List<List<SiteInst>> packDSPCascades(List<List<EDIFHierCellInst>> EDIFDSPCascades) throws IOException {
    List<List<SiteInst>> siteInstCascades = new ArrayList<>();
    writer.write("\n\nPacking DSP Cascades... (" + EDIFDSPCascades.size() + ")");
    for (List<EDIFHierCellInst> cascade : EDIFDSPCascades) {
        List<SiteInst> siteInstCascade = new ArrayList<>();
        // each DSP tile has 2 DSP sites on the Zynq 7000
        writer.write("\n\tCascade Size: (" + cascade.size() + "), Cascade Anchor: "
                + cascade.get(0).getFullHierarchicalInstName());
        // Select a Site location on the device. 
        Site anchorSite = selectDSPAnchorSite(cascade.size());
        SiteTypeEnum siteType = SiteTypeEnum.DSP48E1;
        for (int i = 0; i < cascade.size(); i++) {
            writer.write("\n\t\tPlacing DSP: " + cascade.get(i).getFullHierarchicalInstName());
            Site site = (i == 0) ? anchorSite
                    : device.getSite("DSP48_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
            // Create the SiteInst object
            SiteInst si = new SiteInst(cascade.get(i).getFullHierarchicalInstName(), design, siteType, site);
            // Populate the SiteInst with a DSP Cell. 
            si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
            occupiedSites.get(siteType).add(site);
            availableSites.get(siteType).remove(site);
            si.routeSite();
            siteInstCascade.add(si);
        }
        siteInstCascades.add(siteInstCascade);
    }
    return siteInstCascades;
} // end packDSPCascades()
\end{lstlisting}

\begin{lstlisting}[language=java, caption={Packing LUTGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };

private List<SiteInst> packLUTGroups(List<List<EDIFHierCellInst>> LUTGroups) throws IOException {
    List<SiteInst> LUTSiteInsts = new ArrayList<>();
    writer.write("\n\nPacking LUT Groups...");
    for (List<EDIFHierCellInst> group : LUTGroups) {
        Site selectedSite = selectCLBSite();
        SiteInst si = new SiteInst(group.get(0).getFullHierarchicalInstName(), design,
                selectedSite.getSiteTypeEnum(), selectedSite);
        for (int i = 0; i < group.size(); i++) {
            si.createCell(group.get(i), si.getBEL(LUT6_BELS[i]));
        }
        si.routeSite();
        // for whatever reason, this does not activate DUSED PIP for
        si.addSitePIP(si.getSitePIP("DUSED", "0"));
        si.addSitePIP(si.getSitePIP("CUSED", "0"));
        si.addSitePIP(si.getSitePIP("BUSED", "0"));
        si.addSitePIP(si.getSitePIP("AUSED", "0"));
        LUTSiteInsts.add(si);
    }
    return LUTSiteInsts;
} // end packLUTGroups()
\end{lstlisting}
\begin{multicols}{2}

\subsection{Placement}
    \label{subsec:placement}
    Up until now we have only organized the logical \texttt{EDIFCellInst}s into \texttt{SiteInst}s. 
    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.
